---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "The Cost of Coal production in Pennsylvania on Community Health"
subtitle: "https://github.com/arielocallaghan/V2_Cipparone_Niva_OCallaghan_EDA_Finalproject"
author: "Hugh Cipparone, Katryna Niva and Ariel O'Callaghan"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

getwd()

# Load your packages
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(readxl)
library(imputeTS)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(latticeExtra)
library(tidyverse)
library(lubridate)
library(knitr) #needed for report compilation 
library(tidyverse)
#spatial library additions 
library(sf)
library(leaflet)
library(mapview)
library(leaflet)
library(RColorBrewer)
library(leafsync)
library(lattice)

# Set your ggplot theme

mytheme <- theme_classic(base_size = 14) + theme(axis.text = element_text(color = "black"), legend.position = "top")

theme_set(mytheme)

# Load your datasets


```
# Reference for rmd file report
Knitting commands in code chunks:
  -  `include = FALSE` - code is run, but neither code nor results appear in knitted file
  -  `echo = FALSE` - code not included in knitted file, but results are
  -  `eval = FALSE` - code is not run in the knitted file
  -  `message = FALSE` - messages do not appear in knitted file
  -  `warning = FALSE` - warnings do not appear...
  -  `fig.cap = "..."` - adds a caption to graphical results
** creates bold text
# Header 1
## Header 2
- End a line with two spaces to start a new paragraph.


#Introduction






# Rationale and Research Questions
Write 1-2 paragraph(s) detailing the rationale for your study. This should include both the context of the topic as well as a rationale for your choice of dataset (reason for location, variables, etc.). You may choose to include citations if you like (optional).

At the end of your rationale, introduce a numbered list of your questions (or an overarching question and sub-questions). 
#coal
```{r echo=FALSE}

```

#acs
```{r echo=FALSE}

```



\newpage

#linear regression
```{r}

```

\newpage

# Dataset Information
Provide information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the metadata file for the dataset but formatted in a way that is more narrative.

Describe how you wrangled your dataset in a format similar to a methods section of a journal article.

Add a table that summarizes your data structure (variables, units, ranges and/or central tendencies, data source if multiple are used, etc.). This table can be made in markdown text or inserted as a `kable` function in an R chunk. If the latter, do not include the code used to generate your table.
```{r}

```


\newpage

# Exploratory Analysis 

```{r echo = FALSE}
#code should run but not appear in results 
```


\newpage

# Analysis

#Community Health Indicators
The community health data utilized in this analysis has two separate indicators. One is health outcomes representing the county’s health at that year by looking at many different indicators, including Length of life and quality of life. It weighs both values at 50%. Length of life indicators includes mortality, life expectancy, and premature death. Quality of life indicators includes physical and mental health and low birth weight. The Heath factor is a composite of indicators of health behaviors, clinical care, social and economic factors, and physical environment. (https://www.countyhealthrankings.org/explore-health-rankings/county-health-rankings-model) For this analysis, we utilized Z scores of Health outcomes and Health Factors and the ranking of these variables by county. The data range spans from 2010-2020. county with a ranking of 1 is the best health, and the lowest ranking is the poorest health. A higher Z score indicates worse health for health outcomes and factors. 

A spatial analysis was utilized to display health outcomes rank and Z score by county for 2019 and 2010. 

##Commyunity health data wrangling
```{r echo=FALSE, spatial analysis}
#Disable on-the-fly projections - code from lab sheet 
sf::sf_use_s2(FALSE)

#read in spatial data frame and filter for Pennsylvania 
counties_sf <- st_read('./RawData/cb_2018_us_county_20m.shp')%>% 
  filter(STATEFP == 42) #42|PA|Pennsylvania|01779798

#check data frame information 
colnames(counties_sf)
st_crs(counties_sf) 
nrow(counties_sf)

#Reveal the extent of this dataset via the st_bbox() function
st_bbox(counties_sf)

#View the data
head(counties_sf)

#Plot the county data, colored by area of land in each county
mapView(counties_sf, zcol = "ALAND")

#read in health data frame to join with spatial datframe 
healthdf<- read.csv("./ProcessedData/healthdf.csv",stringsAsFactors = TRUE)

counties_sf$NAME<- as.character(counties_sf$NAME)
class(counties_sf$NAME)

healthdf$County<- as.character(healthdf$County)
class(healthdf$County)

(healthdf)

#counties_sf<- counties_sf %>% 
 # rename(NAME== County)
  
#join data sets together 
counties_sf_join <-  merge(x = counties_sf,
                           y = healthdf, 
                           by.x = "NAME", 
                           by.y = "County" )

#add in coal to look at counties that overlap
coal<- read.csv("./ProcessedData/Coal.Clean.csv",stringsAsFactors = TRUE)
view(coal$County)

class(counties_sf_join$NAME)
NCOL(counties_sf_join$NAME)

#filter for 2019 Data
Health_Data_2019<-counties_sf_join%>%
  filter(year==2019)

#filter for 2010 Data
Health_Data_2010<-counties_sf_join%>%
  filter(year==2010)

#subset coal counties to create new spatial data frame
coal.filter<-counties_sf_join%>% 
  filter(NAME %in% c("Armstrong","Beaver","Bedford","Carbon","Dauphin","Fayette","Greene","Allegheny","Bedfor","Blair","Cambria","Centre","Clarion","Clearfield","Clinton","Columbia","Elk","Indiana","Jefferson","Lackawanna","Lawrence","Luzerne","Lycoming","Mercer","Monongalia","Northumberland","Schuylkill","Sullivan","Tioga"))


```

##Health data visualizations
```{r}
#Visualizations 

#2019 Z score
all.Z.2019<-mapview(Health_Data_2019, 
        zcol = 'Health.Outcomes.Z.Score', 
        col.regions = brewer.pal(10,'OrRd'))

#coal counties only 2019
coal.z.2019<-mapview(coal.filter,
        zcol = 'Health.Outcomes.Z.Score', 
        col.regions = brewer.pal(10,'OrRd'))

sync(all.Z.2019,coal.z.2019)  

#2019 Rank
all.rank.2019<-mapview(Health_Data_2019, 
        zcol = 'Health.Outcomes.Rank', 
        col.regions = brewer.pal(20, 'Blues'))

coal.rank.2019<-mapview(coal.filter, 
        zcol = 'Health.Outcomes.Rank', 
        col.regions = brewer.pal(20, 'Blues'))       

sync(all.rank.2019, coal.rank.2019)      

#2010 Z score
all.z.2010<-mapview(Health_Data_2010, 
        zcol = 'Health.Outcomes.Z.Score', 
        col.regions = brewer.pal(5, 'OrRd'))

coal.z.2010<-mapview(coal.filter, 
        zcol = 'Health.Outcomes.Z.Score', 
        col.regions = brewer.pal(5, 'OrRd'))

#showing maps side by side
sync(all.z.2010, coal.z.2010)

#2010 Rank
all.rank.2010<-mapview(Health_Data_2010, 
        zcol = 'Health.Outcomes.Rank', 
        col.regions = brewer.pal(9, 'Blues'))

coal.rank.2010<-coal.rank.2010<-mapview(coal.filter, 
        zcol = 'Health.Outcomes.Rank', 
        col.regions = brewer.pal(9, 'Blues'))

sync(all.rank.2010, coal.rank.2010)
```

#Time Series Analysis
Although, the length of the health dataset limits time series analysis to a smaller span of years, this is still an important angle of analysis. While the previous linear regression collapsed the data to one representative temporal point, here we seek to focus in on single counties over the time frame allowed by the health data set (2010 until 2020). This will allow the investigation of a more precise match between increased coal output and decreasing community health over time. It is an imprecise look however, because many of the contamination concerns occur in the years following coal mining (water contamination from acid mine drainage). The first step in this process was the wrangling of the available datasets into a comprehensive table.

##Data Wrangling
```{r data wrangling}

#First, we read the previously compiled data sets into the program from where they were saved as csv files
aggCoalData <- read.csv("ProcessedData/aggCoalData.csv")
health.data <- read.csv("ProcessedData/healthdf.csv")

#Then, we had to rename the year column within the health data frame so that the two datasets can be correctly merged together by that column
health.data <- health.data %>%
  rename("Year" = "year")

#Then, we merged the two files together such that the county and year collums match up correctly and we also select for the columns that we will be focusing on for the time series analysis
full.time.series.data <- merge(health.data, aggCoalData, by = c("County", "Year")) %>%
  select(Year, County, Health.Outcomes.Z.Score, Health.Factors.Z.Score, sum.coal.production)

```

##Data Analysis

Now that we have a comprehensive data set to work with that has both the health and coal data by year and by county, we can begin our analysis. To do this, we will be looking at different counties individually. When examined all together, the considerable variability across counties causes high error margins to obscure any observable trends. We will examine any relationship between both health factors and health outcomes z score and coal production. The first county we looked at was Schuylkill County. This county holds the most weight as the peak coal production far exceeded that seen in any other county (can be seen as the highest arch in the plot within the coal data set analysis above). 
```{r Schuylkill County}

schuylkill.data <- filter(full.time.series.data, actual.full.data$County == "Schuylkill")

#comparing health factors z score with coal production in schuylkill county
x <- data.frame(Year = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020))
Year <- schuylkill.data$year
var1<- schuylkill.data$Health.Factors.Z.Score
var2 <- schuylkill.data$sum.coal.production
data <- data.frame(x,var1,var2)

obj1 <- xyplot(var1 ~ Year, data, type = "l" , lwd=2, col="steelblue", ylab = "Health Factors Z Score (Blue)", col.axis = "red")
obj2 <- xyplot(var2 ~ Year, data, type = "l", lwd=2, col="red", ylab = "Coal Produced in Tons (Red)")

doubleYScale(obj1, obj2, add.ylab2 = TRUE, use.style=FALSE)

#linear regression for health factors z score and coal production

schuylkill.health.factors.regression <- lm(data = schuylkill.data, Health.Factors.Z.Score ~ sum.coal.production)

summary(schuylkill.health.factors.regression)

#comparing health outcomes z score in schuylkill county
var3<- schuylkill.data$Health.Outcomes.Z.Score
data <- data.frame(x,var3,var2)

obj1 <- xyplot(var3 ~ Year, data, type = "l" , lwd=2, col="steelblue", ylab = "Health Outcomes Z Score (Blue)", col.axis = "red")
obj2 <- xyplot(var2 ~ Year, data, type = "l", lwd=2, col="red", ylab = "Coal Produced in Tons (Red)")

doubleYScale(obj1, obj2, add.ylab2 = TRUE, use.style=FALSE)

#linear regression for health outcomes z score and coal production

schuylkill.health.outcomes.regression <- lm(data = schuylkill.data, Health.Outcomes.Z.Score ~ sum.coal.production)

summary(schuylkill.health.outcomes.regression)

```
The correlation between the health factors z score and coal production over time in Schuylkill county was the most compelling of the time series analysis. There is a clear correlation visible in the line graph where decreasing coal production is clearly associated with decreasing health factors z score (an indicator of increasing public health). The linear regression completed after this revealed a p-value of 0.000856, far lower than 0.05, indicating a strong correlation. Since we are limited to only 11 data points, this could be coincidental, but considering the extreme output of coal from this region, this may be magnifying any adverse effects of coal production and making this trend more visible. The correlation between health outcomes and coal production was less visible. With a p-value of 0.21, there is likely some correlation, but it is still considerably higher than 0.05. 

We then examined Armstrong county and Clearfield County as two other counties with considerably high coal production to understand how pervasive the correlations previously identified are. 
```{r Armstrong County}

Armstrong.data <- filter(full.time.series.data, actual.full.data$County == "Armstrong")

#comparing health factors z score with coal production in schuylkill county
x <- data.frame(Year = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020))
Year <- Armstrong.data$year
var1<- Armstrong.data$Health.Factors.Z.Score
var2 <- Armstrong.data$sum.coal.production
data <- data.frame(x,var1,var2)

obj1 <- xyplot(var1 ~ Year, data, type = "l" , lwd=2, col="steelblue", ylab = "Health Factors Z Score (Blue)", col.axis = "red")
obj2 <- xyplot(var2 ~ Year, data, type = "l", lwd=2, col="red", ylab = "Coal Produced in Tons (Red)")

doubleYScale(obj1, obj2, add.ylab2 = TRUE, use.style=FALSE)

#linear regression for health factors z score and coal production

armstrong.health.factors.regression <- lm(data = Armstrong.data, Health.Factors.Z.Score ~ sum.coal.production)

summary(armstrong.health.factors.regression)

#comparing health outcomes z score in schuylkill county
var3<- Armstrong.data$Health.Outcomes.Z.Score
data <- data.frame(x,var3,var2)

obj1 <- xyplot(var3 ~ Year, data, type = "l" , lwd=2, col="steelblue", ylab = "Health Outcomes Z Score (Blue)", col.axis = "red")
obj2 <- xyplot(var2 ~ Year, data, type = "l", lwd=2, col="red", ylab = "Coal Produced in Tons (Red)")

doubleYScale(obj1, obj2, add.ylab2 = TRUE, use.style=FALSE)

#linear regression for health outcomes z score and coal production

armstrong.health.outcomes.regression <- lm(data = Armstrong.data, Health.Outcomes.Z.Score ~ sum.coal.production)

summary(armstrong.health.outcomes.regression)

```


```{r Clearfield County}

Clearfield.data <- filter(full.time.series.data, actual.full.data$County == "Clearfield")

#comparing health factors z score with coal production in schuylkill county
x <- data.frame(Year = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020))
Year <- Clearfield.data$year
var1<- Clearfield.data$Health.Factors.Z.Score
var2 <- Clearfield.data$sum.coal.production
data <- data.frame(x,var1,var2)

obj1 <- xyplot(var1 ~ Year, data, type = "l" , lwd=2, col="steelblue", ylab = "Health Factors Z Score (Blue)", col.axis = "red")
obj2 <- xyplot(var2 ~ Year, data, type = "l", lwd=2, col="red", ylab = "Coal Produced in Tons (Red)")

doubleYScale(obj1, obj2, add.ylab2 = TRUE, use.style=FALSE)

#linear regression for health factors z score and coal production

clearfield.health.factors.regression <- lm(data = Clearfield.data, Health.Factors.Z.Score ~ sum.coal.production)

summary(clearfield.health.factors.regression)

#comparing health outcomes z score in schuylkill county
var3<- Clearfield.data$Health.Outcomes.Z.Score
data <- data.frame(x,var3,var2)

obj1 <- xyplot(var3 ~ Year, data, type = "l" , lwd=2, col="steelblue", ylab = "Health Outcomes Z Score (Blue)", col.axis = "red")
obj2 <- xyplot(var2 ~ Year, data, type = "l", lwd=2, col="red", ylab = "Coal Produced in Tons (Red)")

doubleYScale(obj1, obj2, add.ylab2 = TRUE, use.style=FALSE)

#linear regression for health outcomes z score and coal production

clearfield.health.outcomes.regression <- lm(data = Clearfield.data, Health.Outcomes.Z.Score ~ sum.coal.production)

summary(clearfield.health.outcomes.regression)

```
The trends for these counties were much more inconsistent. None had p-values below or near 0.05. Much of this issue may be the result of only having 11 datapoints and so an increase in health data in the more distant past would be crucial to understanding how strong this correlation is. Sadly, much of the more robust data only began a more rigorous collection more recently and so this is difficult to find. 

```{r}

```


## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
