---
title: "Coal Mining and Water Quality Data"
author: "Katryna Niva"
date: "11/26"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r Getting the working directory}

getwd()
```

```{r Installing and loading packages}

library(dplyr)
library(ggplot2)
library(tidyverse)
library(readxl)
library(imputeTS)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
```


#The Coal Mining Dataset
One of our foundational datasets is the one reporting coal production by county each year. This was collected by the U.S. Energy Information Administration and included a robust annual data set that spanned from 1983 until 2021. This took extensive wrangling to combine and narrow to the desired data points. The data wrangling process is shown here:
```{r Wrangling Coal Data}
#inital uploading of all the coal data for each distinct year -- this could not be expedited with data stripping because there were separate links to downloaded spreadsheets instead of links to another website
coal1983 <- read.csv("RawData/coalpublic1983.csv", stringsAsFactors = TRUE)
coal1984 <- read.csv("RawData/coalpublic1984.csv", stringsAsFactors = TRUE)
coal1985 <- read.csv("RawData/coalpublic1985.csv", stringsAsFactors = TRUE)
coal1986 <- read.csv("RawData/coalpublic1986.csv", stringsAsFactors = TRUE)
coal1987 <- read.csv("RawData/coalpublic1987.csv", stringsAsFactors = TRUE)
coal1988 <- read.csv("RawData/coalpublic1988.csv", stringsAsFactors = TRUE)
coal1989 <- read.csv("RawData/coalpublic1989.csv", stringsAsFactors = TRUE)
coal1990 <- read.csv("RawData/coalpublic1990.csv", stringsAsFactors = TRUE)
coal1991 <- read.csv("RawData/coalpublic1991.csv", stringsAsFactors = TRUE)
coal1992 <- read.csv("RawData/coalpublic1992.csv", stringsAsFactors = TRUE)
coal1993 <- read.csv("RawData/coalpublic1993.csv", stringsAsFactors = TRUE)
coal1994 <- read.csv("RawData/coalpublic1994.csv", stringsAsFactors = TRUE)
coal1995 <- read.csv("RawData/coalpublic1995.csv", stringsAsFactors = TRUE)
coal1996 <- read.csv("RawData/coalpublic1996.csv", stringsAsFactors = TRUE)
coal1997 <- read.csv("RawData/coalpublic1997.csv", stringsAsFactors = TRUE)
coal1998 <- read.csv("RawData/coalpublic1998.csv", stringsAsFactors = TRUE)
coal1999 <- read.csv("RawData/coalpublic1999.csv", stringsAsFactors = TRUE)
coal2000 <- read.csv("RawData/coalpublic2000.csv", stringsAsFactors = TRUE)
coal2001 <- read.csv("RawData/coalpublic2001.csv", stringsAsFactors = TRUE)
coal2002 <- read.csv("RawData/coalpublic2002.csv", stringsAsFactors = TRUE)
coal2003 <- read.csv("RawData/coalpublic2003.csv", stringsAsFactors = TRUE)
coal2004 <- read.csv("RawData/coalpublic2004.csv", stringsAsFactors = TRUE)
coal2005 <- read.csv("RawData/coalpublic2005.csv", stringsAsFactors = TRUE)
coal2006 <- read.csv("RawData/coalpublic2006.csv", stringsAsFactors = TRUE)
coal2007 <- read.csv("RawData/coalpublic2007.csv", stringsAsFactors = TRUE)
coal2008 <- read.csv("RawData/coalpublic2008.csv", stringsAsFactors = TRUE)
coal2009 <- read.csv("RawData/coalpublic2009.csv", stringsAsFactors = TRUE)
coal2010 <- read.csv("RawData/coalpublic2010.csv", stringsAsFactors = TRUE)
coal2011 <- read.csv("RawData/coalpublic2011.csv", stringsAsFactors = TRUE)
coal2012 <- read.csv("RawData/coalpublic2012.csv", stringsAsFactors = TRUE)
coal2013 <- read.csv("RawData/coalpublic2013.csv", stringsAsFactors = TRUE)
coal2014 <- read.csv("RawData/coalpublic2014.csv", stringsAsFactors = TRUE)
coal2015 <- read.csv("RawData/coalpublic2015.csv", stringsAsFactors = TRUE)
coal2016 <- read.csv("RawData/coalpublic2016.csv", stringsAsFactors = TRUE)
coal2017 <- read.csv("RawData/coalpublic2017.csv", stringsAsFactors = TRUE)
coal2018 <- read.csv("RawData/coalpublic2018.csv", stringsAsFactors = TRUE)
coal2019 <- read.csv("RawData/coalpublic2019.csv", stringsAsFactors = TRUE)
coal2020 <- read.csv("RawData/coalpublic2020.csv", stringsAsFactors = TRUE)
coal2021 <- read.csv("RawData/coalpublic2021.csv", stringsAsFactors = TRUE)

#inconsistencies in the class of certain columns needed to be resolved in order for the datasets to be properly merged together
coal1996$MSHA.ID <- as.factor(coal1996$MSHA.ID)
coal1997$MSHA.ID <- as.factor(coal1997$MSHA.ID)
coal1998$MSHA.ID <- as.factor(coal1998$MSHA.ID)
coal1999$MSHA.ID <- as.factor(coal1999$MSHA.ID)
coal2000$MSHA.ID <- as.factor(coal2000$MSHA.ID)
coal2001$MSHA.ID <- as.factor(coal2001$MSHA.ID)
coal2002$MSHA.ID <- as.factor(coal2002$MSHA.ID)
coal2003$MSHA.ID <- as.factor(coal2003$MSHA.ID)
coal2004$MSHA.ID <- as.factor(coal2004$MSHA.ID)
coal2005$MSHA.ID <- as.factor(coal2005$MSHA.ID)
coal2006$MSHA.ID <- as.factor(coal2006$MSHA.ID)
coal2007$MSHA.ID <- as.factor(coal2007$MSHA.ID)
coal2008$MSHA.ID <- as.factor(coal2008$MSHA.ID)
coal2009$MSHA.ID <- as.factor(coal2009$MSHA.ID)
coal2010$MSHA.ID <- as.factor(coal2010$MSHA.ID)
coal2011$MSHA.ID <- as.factor(coal2011$MSHA.ID)
coal2012$MSHA.ID <- as.factor(coal2012$MSHA.ID)
coal2013$MSHA.ID <- as.factor(coal2013$MSHA.ID)
coal2014$MSHA.ID <- as.factor(coal2014$MSHA.ID)
coal2015$MSHA.ID <- as.factor(coal2015$MSHA.ID)
coal2016$MSHA.ID <- as.factor(coal2016$MSHA.ID)
coal2017$MSHA.ID <- as.factor(coal2017$MSHA.ID)
coal2018$MSHA.ID <- as.factor(coal2018$MSHA.ID)
coal2019$MSHA.ID <- as.factor(coal2019$MSHA.ID)
coal2020$MSHA.ID <- as.factor(coal2020$MSHA.ID)
coal2021$MSHA.ID <- as.factor(coal2021$MSHA.ID)
coal2002$Average.Employees <- as.factor(coal2002$Average.Employees)
coal2003$Average.Employees <- as.factor(coal2003$Average.Employees)
coal2004$Average.Employees <- as.factor(coal2004$Average.Employees)
coal2005$Average.Employees <- as.factor(coal2005$Average.Employees)
coal2006$Average.Employees <- as.factor(coal2006$Average.Employees)
coal2007$Average.Employees <- as.factor(coal2007$Average.Employees)
coal2008$Average.Employees <- as.factor(coal2008$Average.Employees)
coal2009$Average.Employees <- as.factor(coal2009$Average.Employees)
coal2010$Average.Employees <- as.factor(coal2010$Average.Employees)
coal2011$Average.Employees <- as.factor(coal2011$Average.Employees)
coal2012$Average.Employees <- as.factor(coal2012$Average.Employees)
coal2013$Average.Employees <- as.factor(coal2013$Average.Employees)
coal2014$Average.Employees <- as.factor(coal2014$Average.Employees)
coal2015$Average.Employees <- as.factor(coal2015$Average.Employees)
coal2016$Average.Employees <- as.factor(coal2016$Average.Employees)
coal2017$Average.Employees <- as.factor(coal2017$Average.Employees)
coal2018$Average.Employees <- as.factor(coal2018$Average.Employees)
coal2019$Average.Employees <- as.factor(coal2019$Average.Employees)
coal2020$Average.Employees <- as.factor(coal2020$Average.Employees)
coal2021$Average.Employees <- as.factor(coal2021$Average.Employees)

#The datasets from each year were then merged together into one combined dataset
AllCoal <- full_join(coal1983, coal1984)
AllCoal <- full_join(AllCoal, coal1985)
AllCoal <- full_join(AllCoal, coal1986)
AllCoal <- full_join(AllCoal, coal1987)
AllCoal <- full_join(AllCoal, coal1988)
AllCoal <- full_join(AllCoal, coal1989)
AllCoal <- full_join(AllCoal, coal1990)
AllCoal <- full_join(AllCoal, coal1991)
AllCoal <- full_join(AllCoal, coal1992)
AllCoal <- full_join(AllCoal, coal1993)
AllCoal <- full_join(AllCoal, coal1994)
AllCoal <- full_join(AllCoal, coal1995)
AllCoal <- full_join(AllCoal, coal1996)
AllCoal <- full_join(AllCoal, coal1997)
AllCoal <- full_join(AllCoal, coal1998)
AllCoal <- full_join(AllCoal, coal1999)
AllCoal <- full_join(AllCoal, coal2000)
AllCoal <- full_join(AllCoal, coal2001)
AllCoal <- full_join(AllCoal, coal2002)
AllCoal <- full_join(AllCoal, coal2003)
AllCoal <- full_join(AllCoal, coal2004)
AllCoal <- full_join(AllCoal, coal2005)
AllCoal <- full_join(AllCoal, coal2006)
AllCoal <- full_join(AllCoal, coal2007)
AllCoal <- full_join(AllCoal, coal2008)
AllCoal <- full_join(AllCoal, coal2009)
AllCoal <- full_join(AllCoal, coal2010)
AllCoal <- full_join(AllCoal, coal2011)
AllCoal <- full_join(AllCoal, coal2012)
AllCoal <- full_join(AllCoal, coal2013)
AllCoal <- full_join(AllCoal, coal2014)
AllCoal <- full_join(AllCoal, coal2015)
AllCoal <- full_join(AllCoal, coal2016)
AllCoal <- full_join(AllCoal, coal2017)
AllCoal <- full_join(AllCoal, coal2018)
AllCoal <- full_join(AllCoal, coal2019)
AllCoal <- full_join(AllCoal, coal2020)
AllCoal <- full_join(AllCoal, coal2021)

#This combined dataset was then filtered so that it only included data for the state of Pennsylvania as that is the focus of this inquiry
AllCoal <- filter(AllCoal, Mine.State %in% c("Pennsylvania (Bituminous)", "Pennsylvania (Anthracite)"))

#This data was then slimmed down to the necessary data and the columns were appropriately titled 
AllCoalSlim <- data_frame(AllCoal$Year, AllCoal$Mine.County, AllCoal$Mine.Status, AllCoal$Production..short.tons.)%>%
  rename("Year" = "AllCoal$Year", "County" = "AllCoal$Mine.County", "Status" = "AllCoal$Mine.Status", "Production(tons)" = "AllCoal$Production..short.tons.")

#The coal production was then converted to class integer so that different mines from the same county can be summed together
AllCoalSlim$`Production(tons)` <- as.integer(AllCoalSlim$`Production(tons)`)

#This summation then occured so that there is one sum coal production value for each county each year
aggCoalSlim <- AllCoalSlim %>% group_by(County, Year) %>%
  summarise(sum.coal.production = sum(`Production(tons)`), .groups = 'drop')

#A new column that dictated the weighted average year of production was then calculated in order to have another metric for understanding when the majority of the coal production was happening and what regulations may have been in place at that point
NewCoalSlim <- aggCoalSlim %>%
  pivot_wider(names_from = Year, values_from = sum.coal.production)

NewCoalSlim[is.na(NewCoalSlim)] <- 0

NewCoalSlim <- NewCoalSlim %>% mutate("Year.Max.Production" = (((1983*`1983`)+(1984*`1984`)+(1985*`1985`)+(1986*`1986`)+(1987*`1987`)+(1988*`1988`)+(1989*`1989`)+(1990*`1990`)+(1991*`1991`)+(1992*`1992`)+(1993*`1993`)+(1994*`1994`)+(1995*`1995`)+(1996*`1996`)+(1997*`1997`)+(1998*`1998`)+(1999*`1999`)+(2000*`2000`)+(2001*`2001`)+(2002*`2002`)+(2003*`2003`)+(2004*`2004`)+(2005*`2005`)+(2006*`2006`)+(2007*`2007`)+(2008*`2008`)+(2009*`2009`)+(2010*`2010`)+(2011*`2011`)+(2012*`2012`)+(2013*`2013`)+(2014*`2014`)+(2015*`2015`)+(2016*`2016`)+(2017*`2017`)+(2018*`2018`)+(2019*`2019`)+(2020*`2020`)+(2021*`2021`))/(`1983`+`1984`+`1985`+`1986`+`1987`+`1988`+`1989`+`1990`+`1991`+`1992`+`1993`+`1994`+`1995`+`1996`+`1997`+`1998`+`1999`+`2000`+`2001`+`2002`+`2003`+`2004`+`2005`+`2006`+`2007`+`2008`+`2009`+`2010`+`2011`+`2012`+`2013`+`2014`+`2015`+`2016`+`2017`+`2018`+`2019`+`2020`+`2021`)))

#An additional column is then added to know the total coal production across all years in each county
NewCoalSlim <- NewCoalSlim %>% mutate("sum.coal.prod" = (`1983`+`1984`+`1985`+`1986`+`1987`+`1988`+`1989`+`1990`+`1991`+`1992`+`1993`+`1994`+`1995`+`1996`+`1997`+`1998`+`1999`+`2000`+`2001`+`2002`+`2003`+`2004`+`2005`+`2006`+`2007`+`2008`+`2009`+`2010`+`2011`+`2012`+`2013`+`2014`+`2015`+`2016`+`2017`+`2018`+`2019`+`2020`+`2021`))

#This new metric was then merged with the previous data and then appropriately titled
aggCoalSlim <- left_join(aggCoalSlim, NewCoalSlim, by = "County") 

aggCoalSlim2 <- data_frame(aggCoalSlim$County, aggCoalSlim$Year, aggCoalSlim$sum.coal.production, aggCoalSlim$Year.Max.Production, aggCoalSlim$sum.coal.prod) %>%
  rename("Year" = "aggCoalSlim$Year", "County" = "aggCoalSlim$County", "coal.prod.by.year" = "aggCoalSlim$sum.coal.production", "year.peak.prod" = "aggCoalSlim$Year.Max.Production", "sum.coal.prod" = "aggCoalSlim$sum.coal.prod")

#We now have a core dataset that can be uploaded to the processed data folder for use in the analysis portion. This dataset has the coal production per county per year as well as the summed total coal production and the weighted average year of production for each county
write.csv(aggCoalSlim, file = "./ProcessedData/aggCoalData.csv")

```

Now that the coal data is effectively processed into a comprehensive dataset, we can take an initial look at what data we have. First, we can make a summary table. This will supply the range, median, and interquantile range for each of the parameters included. We can then do some preliminary visualizations. 
```{r Summarizing the Coal Data}
#We can make the summary table of the data within the dataset
coal.data.summary<-aggCoalSlim2 %>% 
  summarize(
    year.range= range(Year),
    annual.coal.production.range= range(coal.prod.by.year),
    total.coal.production.range = range(sum.coal.prod),
    weighted.average.year.range= range(year.peak.prod),
    annual.coal.production.med = median(coal.prod.by.year),
    total.coal.production.med = median(sum.coal.prod),
    weighted.average.year.med = median(year.peak.prod),
    annual.coal.production.iqr = IQR(coal.prod.by.year),
    total.coal.production.iqr = IQR(sum.coal.prod),
    weighted.average.year.iqr = IQR(year.peak.prod),
  )
  
print(coal.data.summary)

#We can also look at the dimensions of the dataset
dim(aggCoalSlim2)

#And the column names to make sure they were labeled correctly
colnames(aggCoalSlim2)

#And make an initial graph of coal production over time in each county
CoalOverTime <- ggplot(aggCoalSlim, aes(`Year`, `sum.coal.production`, group = `County`, color = `County`))+
  geom_smooth()+
  ylab("Total Coal Production (Tons)")+
  ggtitle("Coal Production over Time in each Pennsylvania County")

print(CoalOverTime)

```
The plot over time shows how the coal production has changed over time from 1983 until today in each county. The more recent petering off of most of the mines makes sense as most of the available coal has been removed and the US energy sources have moved away from coal. There is a pretty strong bell curve in most mines which would align with the location of a good mine, the mass export of the mined coal, and then the decrease in production as the most readily available coal is gone. It is interesting to see how all the counties seem to steeply increase in coal production from 1983 on. This could be a limitation in the dataset (only giving strong coverage of mines beginning operation in 1983) but none of that was explained or visible in the metadata and so it is more likely that there was an uptick in coal production at that point. This could be due to the implementation of longwall mining which first began in the 1960s (https://www.dep.pa.gov/Business/Land/Mining/Pages/PA-Mining-History.aspx). It is also clearly visible here that Schuylkill County has far more coal production than any other county.








```{r Installing and loading packages}

#import water quality data sets
WaterQuality <- read.csv("RawData/water_quality_data.csv", stringsAsFactors = TRUE)
WaterQualitySites <- read.csv("RawData/water_quality_location_data.csv", stringsAsFactors = TRUE)
WaterQuality$ActivityStartDate <- mdy(WaterQuality$ActivityStartDate)

#merge the site data with the sampling results
WaterQuality <- left_join(WaterQuality, WaterQualitySites, by = "MonitoringLocationIdentifier")

#filter for pennsylvania
WaterQuality <- filter(WaterQuality, StateCode %in% c(42))

WaterQualitySlim <- data_frame(WaterQuality$ActivityStartDate, WaterQuality$MonitoringLocationName.y, WaterQuality$LatitudeMeasure, WaterQuality$LongitudeMeasure, WaterQuality$CountyCode, WaterQuality$CharacteristicName, WaterQuality$ResultSampleFractionText, WaterQuality$ResultMeasureValue, WaterQuality$ResultMeasure.MeasureUnitCode)%>%
  rename("Date" = "WaterQuality$ActivityStartDate", "Location" = "WaterQuality$MonitoringLocationName.y", "Latitude" = "WaterQuality$LatitudeMeasure", "Longitude" = "WaterQuality$LongitudeMeasure", "County.Code" = "WaterQuality$CountyCode", "Contaminant" = "WaterQuality$CharacteristicName", "State" = "WaterQuality$ResultSampleFractionText", "Measure" = "WaterQuality$ResultMeasureValue", "Units" = "WaterQuality$ResultMeasure.MeasureUnitCode")

WaterQualitySlim$Date <- mdy(WaterQualitySlim$Date)

aggWaterQuality <- WaterQualitySlim %>% group_by(Date, Contaminant)%>%
  summarise(mean(Measure), .groups = 'drop')
  

ContaminantOverTime <- ggplot(aggWaterQuality, aes(Date, `mean(Measure)`, group = Contaminant, color = Contaminant))+
  geom_smooth()+
  ylab("Total Contaminant Concentration (qualitative)")+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank() 
        )

print(ContaminantOverTime)

WaterQualitySlim2 <- WaterQualitySlim %>%
  filter(County.Code == 107) %>%
  filter(WaterQualitySlim$State == "Dissolved")

noFeAndMn <- aggWaterQuality %>% filter(Contaminant == c("Sulfate", `Sulfate as SO4`)) 
onlySulfate <- filter(aggWaterQuality, Contaminant == "Sulfate as SO4")

ContaminantOverTime2 <- ggplot(onlySulfate, aes(Date, `mean(Measure)`, group = Contaminant))+
  geom_point() +
  ylab("Sulfate as SO4 Concentration (mg/l)")


print(ContaminantOverTime2)

```