


```{r}
library(tidyverse)
library(corrplot)
library(ggpubr)

```

```{r}

data<-read.csv("./ProcessedData/health.coal.acs.impute.csv")

```

#Select out 2019 data and remove counties with 0s for white population. Join new coal data (no year - just year of max production and total sum production) to acs and health data. 
```{r}
data.edit<-data %>% 
  filter(perc.white!=0) %>% 
  na.omit() %>% 
  select(!X)
```


```{r}
#run MLR
```{r}

acs.coal.health.regression <- lm(data = data.edit, Health.Outcomes.Z.Score ~ totpop + medage + medincome + perc.white +  perc.bachdegree + year.from.peak.2019 + sum.coal.prod)
summary(acs.coal.health.regression)

#Choose a model by AIC in a Stepwise Algorithm
step(acs.coal.health.regression)

acs.coal.health.regression.2<-lm(data = data.edit, Health.Outcomes.Z.Score ~ medage + perc.white + 
    medincome)
summary(acs.coal.health.regression.2)

```

#Check MLR assumptions
```{r}
par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(acs.coal.health.regression.2)
par(mfrow = c(1,1))

#remove county and other non-informative variables to permit correlation plot
acs.health.coal.cor.edit <- data.edit %>% 
  select(!c(County,totpop, totpopwhite, bachdegree,Health.Factors.Z.Score,
            Health.Factors.Rank, year)) %>% 
  na.omit()


acs.health.coal.Corr <- cor(acs.health.coal.cor.edit)
corrplot(acs.health.coal.Corr, method = "circle", type="upper")
corrplot.mixed(acs.health.coal.Corr, upper = "circle")
```

#Visualize scatterplots of all important variables to check for need for transformations and linearity assumption

```{r}

health.age<-ggplot(data.edit, aes(medage,Health.Outcomes.Z.Score))+
  geom_point()+
  ylab("Health Outcomes")+
  xlab("Median Age")+
  theme(axis.text.x = element_text(angle = 90))+
  theme(legend.position="none")+
  theme_classic()

health.race<-ggplot(data.edit, aes(perc.white,Health.Outcomes.Z.Score))+
  geom_point()+
  ylab("Health Outcomes")+
  xlab("% White-Identifying")+
  theme(axis.text.x = element_text(angle = 90))+
  theme(legend.position="none")+
  theme_classic()

health.inc<-ggplot(data.edit, aes(medincome,Health.Outcomes.Z.Score))+
  geom_point()+
  ylab("Health Outcomes")+
  xlab("Median Income (USD)")+
  theme(axis.text.x = element_text(angle = 90))+
  theme(legend.position="none")+
  theme_classic()

ggarrange(health.age, health.race, health.inc,
align='hv', labels=c('A', 'B','C'),
nrows=3)

```

